---
title: "2/24/20 notes soc 613"
author: "JP"
date: "2/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes

#Heteroscedascity
Which factors are modifying the beta parameters?
`\mu_{0j}` and `\epsilon_{0ij}`

epsilon for age `\epsilon_{1ij}` `(age_{ij})`

Consider heteroscedascity at level 1 because it could change the variance in level 2

Panel data is often NOT constant over time.

- tend to see more variance in variables over time

If you're doing a random slopes model, you should probably look at heteroscedascity at level 1. 

```{r}
library(tidyverse)
library(readstata13)
library(lme4)
library(psych)
library(optimx)
library(lmerTest)
library(dfoptim)


options(max.print = 99999)
options(scipen = 999)

getwd()

set.seed(232020)

data <- read.dta13("E:/UO/R Projects/SOC 613/data/AddHealth.dta")
```

`https://quantdev.ssri.psu.edu/sites/qdev/files/ILD_Ch06_2017_MLMwithHeterogeneousVariance.html`


```{r}
model <- lmer(cesd ~ age_w1 + (1 | schoolid), data = data)
summary(model)

# what it should look like
# .mixed cesd age_w1 || schoolid: || aid: age_w1, noconstant

data %>% 
  dplyr::select(age_w1) %>% 
  inspectdf::inspect_na()

mod1 <- nlme::lme(fixed = cesd ~ age_w1 + 1,  
                random = list(schoolid = nlme::pdSymm(form = ~ schoolid)),
                weights = nlme::varIdent(form = ~ 1 + age_w1 | aid),
                        data = data)
summary(mod1)

library(nlme)

ex <- lme(fixed = cesd ~ age_w1, 
                random = ~ 1 | schoolid,
                weights = varIdent(form = ~ age_w1 | aid), 
                data = data)


```

